The Astropy Project

### 2017-11-11 main.tex
remove comment
major edits to the introduction

\section{Introduction} \label{sec:intro}
% first draft by Moritz (hamogu)
All astronomical research makes use of software in some way and
astronomy as a field has long supported the development of specialized
software tools for different astronomical tasks. Some of those
software packages are (or at least were) supported by large
institutions, e.g.\ IRAF \citep[developed at NOAO,][]{IRAF}, MIDAS
\citep[developed at ESO,][]{MIDAS}, or ds9 \citep[developed at
% * <bsipocz@gmail.com> 2017-09-28T21:05:00.569Z:
%
% > or ds9
% probably starlink is a better example as being a large package of software similarly to IRAF and MIDAS rather than a single tool.
%
% ^.
% hamogu: Feel free ot swap in Starlink, if you think that's more appropriate. I picked ds9 becase (a) it's wider known (e.g. in the X-ray community) and (b) because it is an example of a widely used tool with institutional support that is NOT a large package. TOPCAT would have been another example if we want to expand that list.
% olebole: TOPCAT is part of Starlink Java. An other examples would be CASA resp. AIPS. I'd rather omit ds9, since it is mainly developed by one person.
% timj: Starlink originally developed TOPCAT. Starlink is much better an example than DS9.
SAO,][]{ds9}. These packages tend to address a wide range of users
and typically provide some level of documentation and user
support is provided. Other packages are developed foremost by individual research
groups and used mostly in work by members of the group. Those
packages can address more specialized needs, but are not necessarily
available to the community and even if they are, documentation can in
some cases be insufficient for new users and the quality of the
software, i.e.\ the robustness and the level of testing that has been
performed to ensure correctness, is not always clear. On the other
hand, there are also many widely used, well documented, high quality
software packages which are developed by individuals or small groups.

The implementation of astronomical software, be it for a package meant
for wide distribution or scripts and programs for a specific research
project, can be eased by the use of a library that provides core
functionality that is common to many astronomical tasks. Reading and
writing FITS files is one example where it is time
consuming to implement the FITS standard correctly and
programming is much faster when using an existing library to perform this
task. Another example of such a common task is dealing with different
astronomical coordinate systems. The \astropy project aims to provide a
library of core astronomical functionality in the \python programming
language.

\python\footnote{\url{https://www.python.org/}} is an increasingly popular
general-purpose scripting language that is available under a permissive open
source software licence free of charge for all major operating systems. Stable
and well-developed packages provide support for array arithmetic
\citep[\package{numpy},][]{numpy}, a wide variety of functions for scientific
computing \citep[\package{scipy},][]{numpy}, and publication-quality plotting
\citep[\package{matplotlib},][]{matplotlib}. Tens of thousands of other packages
are available which can help with tasks that are not astronomy specific but
might be performed in the course of astronomical research, e.g.\ presenting
results on websites or interfacing with databases.

At the same time, an ecosystem of tools, programs, and web services
has become available that makes it easier to collaborate on code development,
continuously test the code, build and provide documentation, and
supply downloadable packages. The \astropy project aims to develop and
provide high quality code and documentation according to the best
practices in software development. The project makes use of these
tools to do so without central institutional oversight.

The first public release of the \astropy package is described in
\cite{astropy}. Since then, the \astropy package has been
used in hundreds of projects, and the scope of the package has grown
considerably. At the same time, the community of astronomers
contributing to the project has grown tremendously and an ecosystem
of packages supporting or affiliated with the \astropy core has
developed. In this paper, we describe the current status of the
\astropy community, the \astropypkg core package, and the \astropy
ecosystem.

We start by describing the way the \astropy project works and is
organized in \sectionname~\ref{sec:org}.  The project develops a core
package called \astropypkg (\sectionname~\ref{sec:core}) and several
separate packages that help setting up the infrastructure for testing,
documentation etc.\ (\sectionname~\ref{sec:infrastructure}).
We end with a short vision for
the future of \astropy in particular and astronomical software in general
in \sectionname~\ref{sec:future}.

@@

\inlinecomment{DavidS: we'd like to build on this to allow use of a
context manager for selecting a version of constants. Here, or in the Future
section?}

>>>>

\section{Introduction} \label{sec:intro}
% first draft by Moritz (hamogu)
% heavily edited by Adrian PW - let me know if you want to revert anything!
All astronomical research makes use of software in some way.
Astronomy as a field has thus long supported the development of specialized
software tools for astronomical tasks, from software packages that enable
individualized scientific research to data reduction tools and pipelines that
enable collaborations and survey operations.
Some such software packages are or were supported by large institutions and are
intended for a wide range of users.
These packages typically provide some level of documentation and user support or
training.
Other packages are developed by individual researchers or research groups and
are then typically used by smaller groups for more domain-specific purposes.
Whether for a package meant for wide distribution or for scripts and programs
for a specific research project, the implementation of astronomical software can
be eased through the use of a library that provides core functionality that is
common to many astronomical tasks.
The users of such software then also benefit from a community and ecosystem
built around a common foundation.
The \astropy project has grown to become this community, and the \astropypkg
core package a feature-rich library for \python-based astronomical software.

The development of the \astropypkg core package began as a largely
community-driven effort to standardize core functionality for astronomical
software in \python.
In this way, its genesis differs from but builds upon many substantial and
former astronomical software development efforts that were commissioned or
initiated through large institutional support, for example IRAF \citep[developed
at NOAO;][]{IRAF}, MIDAS \citep[developed at ESO;][]{MIDAS}, or Starlink
\citep[originally developed by a consortium of UK institutions and now
maintained by the East Asian Observatory;][]{starlink1982,starlink2013}.
% More recently, community-driven efforts have seen significant success in the astronomical sciences \citep{yt}

The \astropypkg core package is now a feature-rich library of sufficiently
general tools and classes that supports the development of more specialized
code.
An example of such functionality is reading and writing FITS files: it would be
time consuming and impractical for multiple groups to implement the FITS
standard \citep{FITS} and maintain software for such a general-purpose need.
Another example of such a common task is in dealing with representations of and
transformations between astronomical coordinate systems.
The \astropy project aims to provide an open-source, open-development core
package (\astropypkg) and an ecosystem of \emph{affiliated packages} that
support astronomical functionality in the \python programming language.

\python\footnote{\url{https://www.python.org/}} is an increasingly popular,
general-purpose scripting language that is available under a permissive open
source software license free of charge for all major operating systems.
Stable and well-developed packages provide support for array arithmetic
\citep[\package{numpy};][]{numpy}, a wide variety of functions for scientific
computing \citep[\package{scipy};][]{numpy}, and publication-quality plotting
\citep[\package{matplotlib};][]{matplotlib}. Tens of thousands of other packages
are available which can help with tasks that are not astronomy specific but
might be performed in the course of astronomical research, for example interfacing with databases or statistical inference.

\python has become especially popular in the quantitative sciences, especially
where researchers must simultaneously produce research, perform data analysis,
and develop software.
A large part of this success owes itself to a vibrant community of developers
and a continuously-growing ecosystem of tools and web services that enable
easier collaboration on software development, easier writing and sharing
documentation, and continuous testing and validation of software.
More recently, package managers such as Anaconda\url{https://anaconda.org/} have
streamlined the installation process for most packages, therefore significantly
reducing the barrier to entry for using many such libraries.

The \astropy project aims to develop and provide high quality code and
documentation according to the best practices in software development.
The project makes use of these tools to do so without central institutional
oversight.
The first public release of the \astropypkg package is described in
\cite{astropy}. Since then, the \astropypkg package has been
used in hundreds of projects, and the scope of the package has grown
considerably. At the same time, the community of astronomers
contributing to the project has grown tremendously and an ecosystem
of packages supporting or affiliated with the \astropypkg core has
developed.
In this paper, we describe the current status of the \astropy community and the
\astropypkg core package and discuss goals for future development

We start by describing the way the \astropy project functions and is organized
in \sectionname~\ref{sec:org}. We then describe the main software efforts
developed by the \astropy project itself: a core package called \astropypkg
(\sectionname~\ref{sec:core}) and several separate packages that help maintain
the infrastructure for, e.g., testing and documentation
(\sectionname~\ref{sec:infrastructure}). We end with a short vision for
the future of \astropy in particular and astronomical software in general
in \sectionname~\ref{sec:future}.

@@
